---
title: "GEE Modeling"
author: "Abhijit"
date: "1/13/2019"
output:
  word_document: default
  html_document:
    self-contained: no
---

We modeled the relative likelihood of having a syndesmophyte at a particular sector location of a
particular spinal joint using a generalized estimating equations (GEE) model to account for repeated
measurements for each patient. This GEE model modeled the presence of syndesmophytes at each sector
of each spinal joint (dependent variable) on the sector location, spinal joint, distance of the
sector from the aorta (in mm), and the interaction between the sector location and its distance from
the aorta. We used a working exchangeable correlation structure and computed robust standard errors
using sandwich estimators. [need a description of how the sectors are labeled, compared to aorta].
We see that the effect of aorta distance strongly depends on the sector location, even accounting
for the effect of spinal joint. The following table summarizes the odds ratios for aorta distance by
sector location. In each of the following tables, the odds ratios corresponding to aorta distance
reflect the change in risk for 1 mm __decrease__ in distance, i.e. how does risk change when the
aorta is __closer__ to the spine.

```{r, echo = FALSE, message = F, warning = F, results = 'asis'}
ProjTemplate::reload()
datadir <- file.path(ProjTemplate::find_dropbox(), 'NIAMS','Ward','Tan_Aorta')
library(glue)

# m0 <- geeglm(SyndPresent ~ rel_angle, data = munged_data %>% mutate(Dist2Aorta = -Dist2Aorta), # Protective as you get closer
#              id = ID,
#              family = binomial(link=logit),
#              corstr = 'ex',
#              std.err = 'san.se')
# m8 <- update(m0, . ~ . + rel_angle*Dist2Aorta + Spine)
# saveRDS(m8, file = file.path(datadir,'FinalGEEmodel.rds'), compress = T)

m8 <- readRDS(file.path(datadir, 'FinalGEEModel.rds'))
L <- cbind(matrix(0,36, 48), diag(rep(1,36))) # Contrast matrix
L <- rbind(L[1:18,],rep(0,84),L[19:36,])
L[,38] = 1 # Dist2Aorta location

bl <- doBy::esticon(m8, L)
labs <- str_remove(names(coef(m8)[49:84]),':Dist2Aorta')
labs <- c(labs[1:18], 'rel_angle0', labs[19:36])
bl <- cbind('Sector' =labs, bl) %>%
  mutate(Sector = str_remove(Sector, 'rel_angle')) %>%
  select(Sector, Estimate, Std.Error,`Pr(>|X^2|)`, Lower, Upper) %>% 
  rename(pval = `Pr(>|X^2|)`)

bl %>%
  mutate_at(vars(Estimate, Lower, Upper), funs(round(exp(.), 2))) %>%
  mutate(out = as.character(glue_data(., '{Estimate} ({Lower}, {Upper})'))) %>%
  select(Sector, out, pval) %>% 
  mutate(pval = ProjTemplate::myformat_pvalue(pval, digits = 3)) %>% 
  mutate(pval = str_replace(pval, 'e(-\\d{2})', ' x 10^\\1^')) %>% 
  rename(`OR (95% CI)` = out, `P-value` = pval) %>% 
  knitr::kable()
```

The p-values for these odds ratios are shown in the following figure, and shows that the effect of
aorta distance is most statistically significant near the nadir sector and decreases as you move away
from the nadir in either direction; they remain significant up to about 45 degrees from the nadir
sector. 

```{r pvals, echo = F, fig.path = 'graphs/', dev = c('png', 'pdf'), fig.width = 8, fig.height = 4}
scientific_10 <- function(x) {
  parse(text=gsub("1e", "10^", scales::scientific_format()(x)))
}

bl %>% 
  mutate(Sector = as.numeric(Sector)) %>% 
  ggplot(aes(Sector, pval))+geom_point()+
    geom_hline(yintercept = 0.05, linetype =3)+
    scale_y_log10('P-value', label=scientific_10, 
                  breaks = c(0.05, 1e-3, 1e-7, 1e-11))+
    scale_x_continuous('Sector angle') +
    theme_bw()
```

The following plot shows the odds ratio of having a syndesmophyte for every 1 mm the aorta is closer
to the sector location. At the sector closest to the aorta, each mm the aorta is closer, the risk 
of having a syndesmophyte halves. This may indicate that the closer the pulsing is to the spine, 
the lower the risk of syndesmophyte growth. _(This includes 95% CI, next one doesn't)_

```{r OR_CI, echo = F, fig.path = 'graphs/', dev = c('png','pdf'), fig.width=8, fig.height = 4}
bl %>% 
  mutate(Sector = as.numeric(Sector)) %>% 
  mutate(OR = exp(Estimate), 
         Lower = exp(Estimate - 1.96*Std.Error), 
         Upper = exp(Estimate + 1.96*Std.Error)) %>% 
  ggplot(aes(x = Sector, y = OR, ymin = Lower, ymax = Upper))+
    geom_pointrange() +
    geom_hline(yintercept = 1, linetype = 2) +
    scale_y_continuous('Odds ratio') + 
     theme_bw()

```

Alternatively,

```{r OR_noCI, echo = F, fig.path = 'graphs/', dev = c('png','pdf'), fig.width=8, fig.height = 4}
bl %>% 
  mutate(Sector = as.numeric(Sector)) %>% 
  mutate(OR = exp(Estimate), 
         Lower = exp((Estimate - 1.96*Std.Error)), 
         Upper = exp((Estimate + 1.96*Std.Error))) %>% 
  ggplot(aes(x = Sector, y = OR, ymin = Lower, ymax = Upper))+
    geom_point() +
    geom_hline(yintercept = 1, linetype = 2) +
    scale_y_continuous('Odds ratio') + 
     theme_bw()

```


### Supplementary data

The results from fitting the full GEE model is given below:

```{r, echo = F, message = F, warning = F}
library(glue)
m8 <- readRDS(file.path(datadir, 'FinalGEEModel.rds'))
bl2 <- broom::tidy(m8)
bl2 <- bl2 %>%
  mutate(term = str_replace(term, 'rel_angle', 'Sector = ')) %>%
  mutate(term = str_replace(term, 'Spine', 'Spinal joint = ')) %>%
  mutate(term = str_replace(term, '([TL]\\d+)([TL]\\d+)','\\1-\\2')) %>% 
  mutate(term = str_replace(term, 'Dist2Aorta', 'Aorta distance (mm)')) %>%
  mutate(OR = exp(estimate)) %>%
  mutate(CI = as.character(glue_data(.,
    '({round(exp(estimate - 1.96*std.error),2)}, {round(exp(estimate + 1.96*std.error),2)})'
  ))) %>% 
  mutate(p.value = ProjTemplate::myformat_pvalue(p.value, digits = 3)) %>%
  mutate(term = str_replace(term,':', ' x ')) %>%
  select(term, OR, CI, p.value) %>%
  filter(term != '(Intercept)') %>% 
  mutate(p.value = str_replace(p.value, 'e(-\\d{2})', ' x 10^\\1^')) %>% 
  set_names(c('Variable','Odds Ratio','95% CI','P-value'))
knitr::kable(bl2)

```

If we fit a GEE model using the sector angles as continuous, we see that for every 5 degrees we move away from the nadir sector (I've modeled it for absolute angle), the odds of having growth increases 29%, and the
risk also decreases 6% for every mm the aorta is closer to the sector. There is a statistically signficant
sector by aorta distance interaction.

```{r cts, echo = F, message = F, warning = F}
# munged_data <- readRDS(file.path(datadir,'rda','cleaned_data_2.rds'))
# munged_data %<>% mutate( rel_angle = as.factor(rel_angle)) %>%
#   mutate(rel_angle = fct_relevel(rel_angle, '0')) %>%
#   mutate(Spine = fct_relevel(Spine, 'T12L1')) %>%
#   arrange(ID)
# munged_data3 <- munged_data %>% 
#   mutate(rel_angle = abs(as.numeric(rel_angle)/5)) %>% 
#   mutate(Dist2Aorta = -Dist2Aorta)
# m8_1 <- geeglm(formula = m8$formula, data = munged_data3, 
#                family = binomial(link = logit),
#                id = ID,
#                corstr = 'ex',
#                std.err = 'san.se')
# saveRDS(m8_1, file.path(datadir, 'FullGEEcts.rds'))
m8_1 <- readRDS(file.path(datadir, 'FullGEEcts.rds'))
bl3 <- broom::tidy(m8_1)
bl3 <- bl3 %>%
  mutate(term = str_replace(term, 'rel_angle', 'Sector angle (per 5 degrees)')) %>%
  mutate(term = str_replace(term, 'Spine', 'Spinal joint = ')) %>%
  mutate(term = str_replace(term, '([TL]\\d+)([TL]\\d+)','\\1-\\2')) %>% 
  mutate(term = str_replace(term, 'Dist2Aorta', 'Aorta distance (mm)')) %>%
  mutate(OR = round(exp(estimate),2)) %>%
  mutate(CI = as.character(glue_data(.,
    '({round(exp(estimate - 1.96*std.error),2)}, {round(exp(estimate + 1.96*std.error),2)})'
  ))) %>% 
  mutate(p.value = ProjTemplate::myformat_pvalue(p.value, digits = 3)) %>%
  mutate(p.value = str_replace(p.value, 'e(-\\d+)',' x 10^\\1^')) %>% 
  mutate(term = str_replace(term,':', ' x ')) %>%
  filter(term != '(Intercept)') %>% 
  select(term, OR, CI, p.value) %>% 
  set_names(c('Variable','Odds Ratio', '95% CI', 'P-value'))
knitr::kable(bl3)
```

